---
title: "Convergence diagnostics"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PEPI)
library(cmdstanr)
library(bayesplot)
```

`fit_multirates()` defaults to `method = "variational"` (ADVI), a fast
approximation with no notion of chains, R-hat, divergences or treedepth.
Whenever you need to trust a result - not just eyeball it - fit with
`method = "sample"` (full Hamiltonian Monte Carlo / NUTS) instead, and use
the diagnostics below to check the chains actually converged. These are
thin wrappers around [bayesplot](https://mc-stan.org/bayesplot/), which
already knows how to read a `cmdstanr` fit directly, so there is no PEPI-specific
plotting logic to learn.

## Simulate data and fit with MCMC

```{r}

set.seed(42)

sim = simulate_multirates_tree(
  sampling_times = c(3, 6, 9),
  tmrca = 1, t_min = 0,
  lambda_n = 1, s_epi = 0.15,
  omega_n_wt = 5e-3, omega_p_wt = 2e-3,
  drivers = list(
    list(id = "KRAS", t_driver = 2, s_driver = 0.3,
        omega_n_driver = 5e-3, omega_p_driver = 2e-3,
        ms_driver = -0.5, sigma_driver = 0.5,
        alpha_n_driver = 1, beta_n_driver = 10,
        alpha_p_driver = 1, beta_p_driver = 10)
  ),
  clades_wt = list(
    list(id = "c1", t_clade_wt = 1.5)
  ),
  mu = 1e-7, l = 2.7e9, kappa = 20, sigma_count = 0.1,
  seed = 42
)

x = init_multirates(sim$tables, m_trunk = sim$truth$m_trunk)

x = fit_multirates(x, cmdstan_path = cmdstanr::cmdstan_path(),
                   method = "sample", chains = 2, ndraws = 200, seed = 45,
                   mu = 1e-7, l = 2.7e9, t_min = 0, ms_epi = 0, sigma_epi = 0.5,
                   alpha_lambda = 1, beta_lambda = 1,
                   alpha_n_wt = 1, beta_n_wt = 10, alpha_p_wt = 1, beta_p_wt = 10,
                   min_kappa = 5, max_kappa = 200,
                   min_sigma_count = 0.01, max_sigma_count = 1)

```

## One-call dashboard

`plot_diagnostics()` arranges the four checks you should look at first -
R-hat, effective sample size, divergences and the HMC energy diagnostic -
into a single figure.

```{r, fig.width=10, fig.height=8}
plot_diagnostics(x)
```

- **R-hat** (top left): the potential scale reduction factor, one point per
  parameter. Values close to 1 mean the chains agree with each other;
  anything above about 1.05 means they haven't converged to the same
  distribution and you need more iterations (or the model is
  misspecified).
- **Effective sample size ratio** (top right): $N_{eff}/N$, one point per
  parameter. Low ratios mean the chain is highly autocorrelated - the raw
  draws contain less independent information than their count suggests.
- **Divergences** (bottom left): transitions the sampler had to abandon
  because the numerical integrator became unstable. Even a handful can bias
  posterior estimates in the region where they occur.
- **Energy / E-BFMI** (bottom right): the marginal energy distribution
  ($\pi_E$) should closely track the energy transition distribution
  ($\pi_{\Delta E}$); a systematic mismatch signals inefficient exploration.

## Digging deeper into individual diagnostics

Each panel above is also available on its own, and accepts a `params`
argument (Stan parameter base names, no brackets - indexed parameters like
`s_driver` are expanded to all of their elements automatically) to zoom in
on a specific subset.

```{r, fig.width=9, fig.height=6}
plot_trace(x, params = c("lambda_n", "s_epi", "s_driver"))
```

Trace plots overlay each chain's path through the parameter space over
iterations; well-mixed chains look like "fuzzy caterpillars" with no trend
and substantial overlap between chains.

```{r, fig.width=9, fig.height=6}
plot_acf(x, params = c("lambda_n", "s_epi", "s_driver"))
```

Autocorrelation should decay quickly with lag; slow decay is the same
problem the effective-sample-size panel flags, just visualised directly.

```{r, fig.width=7, fig.height=6}
plot_pairs(x, params = c("lambda_n", "s_epi"))
```

Pairs plots are the standard way to spot the funnel-shaped geometries that
cause divergences (marked in red, if any) - a divergence cluster
concentrated in a narrow region of this plot points at exactly which
parameters are hard to sample jointly.

```{r, fig.width=7, fig.height=5}
plot_treedepth(x)
```

Chains that repeatedly hit the maximum treedepth are being cut off before
making a U-turn; this hurts sampling efficiency (though, unlike
divergences, it does not necessarily bias the posterior).

## What to do if something looks wrong

- **High R-hat / low ESS**: run more iterations (`ndraws`), or more chains
  (`chains`) so R-hat has more information to detect disagreement.
- **Divergences**: tighten the priors that are letting the sampler explore
  an implausible, poorly-conditioned region (e.g. narrow `min_kappa`/
  `max_kappa`, `min_sigma_count`/`max_sigma_count` bounds, or gamma priors
  on the switch rates `omega_*` closer to the scale you actually expect).
- **Both persist**: fall back to `method = "variational"` for fast
  iteration while adjusting the model/data, and only switch to
  `method = "sample"` once instructed as a final check.
