|                                                       |
|-------------------------------------------------------|
| title: "Inference of Epigenetic Plasticity with Driver events with PEPI" |
| author: "PEPI Developers"                             |
| date: "`r Sys.Date()`"                                |
| output: html_document                                 |

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Introduction

This vignette illustrates how PEPI can be used to analyze systems in which
selective driver mutations interact with epigenetic state dynamics.

Unlike the other examples, here clonal expansion is driven by
genetic selection, potentially confounding epigenetic inheritance signals.
PEPI explicitly models this interaction in a unified Bayesian framework.
The dataset is simulated with ProCess, a cancer evolution simulator available at https://github.com/caravagnalab/ProCESS. Cancer cells were collected at two time points and sorted according to their epistate. On both subset of cells, WGS is simulated in order to obtain a VAF coordinate for any time point and epistate.

We will:

• visualize the lineage tree and driver placement
• inspect VAF structure
• fit a joint driver–epigenetic model
• interpret posterior estimates

# Load Libraries

```{r setup}
library(PEPI)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(reshape2)
```

# Forest Structure

```{r no-driver-forest}
forest <- PEPI::driver_example$forest
PEPI::plot_forest(forest)
```

The lineage tree shows a dominant expanding branch seeded by a driver event.
Epigenetic states are partially conserved but not perfectly aligned with clonal
structure, indicating interaction between selection and state switching.


## VAF Space

```{r no-driver-vaf}
muts <- PEPI::driver_example$muts

get_branch = function(forest,cell_id,stop_id){
  
  s = cell_id
  branch = s
  
  while(s > stop_id){
    
    s =  forest %>% filter(cell_id == s) %>% pull(ancestor)
    if(s > stop_id){branch = c(branch,s)}
  }
  
  return(branch)
  
}

id_B = forest %>% as_tibble() %>% filter(mutant == "B") %>% pull(cell_id) %>% min()

id_cl = forest %>% as_tibble() %>% filter(mutant == "B",epistate == "+") %>% pull(cell_id) %>% min()

branch_dr = get_branch(forest,cell_id = id_B,stop = 0)
branch_cl = get_branch(forest,cell_id = id_cl,stop = id_B)

muts = muts %>% mutate(label = ifelse(cell_id %in% branch_dr,"driver",
    ifelse(cell_id %in% branch_cl,"first nested clade","subclonal"))) %>%
  mutate(label = ifelse(cell_id == 0,"truncal",label))


p1 <- PEPI::plot_vaf(muts, variables = c("vaf_1_n", "vaf_1_p"))
p2 <- PEPI::plot_vaf(muts, variables = c("vaf_2_n", "vaf_2_p"))
ggarrange(plotlist= list(p1$scatterplots[[1]],p2$scatterplots[[1]]),nrow = 1, ncol = 2,common.legend = T) 
```

Driver-associated mutations form a high-VAF cluster. This cluster reaches a higher VAF in the
$\oplus$ sample because, at the time of driver emergence, the $\oplus$ population is smaller,
and the driver therefore expands while competing with fewer cells. We also observe a distinct
cluster associated with the first nested clade within the driver lineage.



## PEPI object construction

```{r no_driver obj construction}
m_clade = muts %>% filter(label == "first nested clade") %>% nrow()
m_driver = muts %>% filter(label == "driver") %>% nrow()

vaf_clade = muts %>% filter(label == "first nested clade") %>% dplyr::select(starts_with("vaf")) %>% 
  reshape2::melt() %>% group_by(variable) %>% summarize(v = mean(value)) %>% 
  pivot_wider(names_from = variable,values_from = v)

vaf_driver = muts %>% filter(label == "driver",cell_id == id_B) %>% dplyr::select(starts_with("vaf"))%>%  reshape2::melt() %>% filter(value < Inf) %>% 
  group_by(variable) %>% summarize(v = mean(value)) %>% 
  pivot_wider(names_from = variable,values_from = v)


clade_statistics <- tibble(
      cluster_name = c("S1","S2"),
      n_muts = c(m_clade,m_driver),
      vaf_1_n = c(vaf_clade$vaf_1_n,vaf_driver$vaf_1_n),
      vaf_1_p = c(vaf_clade$vaf_1_p,vaf_driver$vaf_1_p),
      vaf_2_n = c(vaf_clade$vaf_2_n,vaf_driver$vaf_2_n),
      vaf_2_p = c(vaf_clade$vaf_2_p,vaf_driver$vaf_2_p),
      is_clade = c(TRUE,FALSE),
      names = c("B","B"),
      driver_type = c("dc","dc"),
      has_driver = c(TRUE,TRUE),
      rewind = c(FALSE,FALSE)
)


times = forest %>% filter(!is.na(sample)) %>% group_by(sample) %>% 
  summarize(t = max(birth_time)) %>% pull(t) %>% sort()

counts = driver_example$counts

n_times = 2

c1 = counts %>% filter(time >= times[1] - 0.01 & time < times[1] + 0.01)
c1 = c1[c(nrow(c1)-1,nrow(c1)),]
c2 = counts %>% filter(time >= times[2] - 0.01 & time < times[2] + 0.01)
c2 = c2[c(nrow(c2)-1,nrow(c2)),]

ztot  = c(c1 %>% pull(count) %>% sum(),c2 %>% pull(count) %>% sum())

rho_1 = forest %>% filter(sample == "Sampling_1") %>% group_by(epistate) %>% summarize(n = length(cell_id))
ccf_1_n = (rho_1 %>% filter(epistate == "-") %>% pull(n))/(rho_1 %>% pull(n) %>% sum())
ccf_1_p = (rho_1 %>% filter(epistate == "+") %>% pull(n))/(rho_1 %>% pull(n) %>% sum())

rho_2 = forest %>% filter(sample == "Sampling_2") %>% group_by(epistate) %>% summarize(n = length(cell_id))
ccf_2_n = (rho_2 %>% filter(epistate == "-") %>% pull(n))/(rho_2 %>% pull(n) %>% sum())
ccf_2_p = (rho_2 %>% filter(epistate == "+") %>% pull(n))/(rho_2 %>% pull(n) %>% sum())

zminus = c(ccf_1_n,ccf_2_n)*ztot
zplus =  c(ccf_1_p,ccf_2_p)*ztot


counts <- tibble(
  count_n = zminus,
  count_p = zplus,
  time = times
)

mu = driver_example$mu$mean
  
genomic_constants <- tibble(
  variable = c("genome_length", "mu"),
  value = c(51e6, mu)
)

pepi_obj = init(clade_statistics = clade_statistics,counts = counts,
                genomic_constants = genomic_constants)

pepi_obj
```

The object requires clade statistics (number of mutations, ccf), total cell counts at the different time points and genomic constants as mutation rate and genomic length.

## Model Fitting

```{r driver-fit}

pepi_obj <- fit_pepi(
  pepi_obj,
  model_type = "logistic",
  ms_driver_n = NULL,
  sigma_driver_n = NULL,
  ms_dc = -2,
  sigma_dc = 1.5,
  ms_cd = NULL,
  sigma_cd = NULL,
  alpha_lambda = 0.6, beta_lambda = 1,
  alpha_plus = 10, beta_plus = 220,
  alpha_minus = 10, beta_minus = 220,
  alpha_n = 10, beta_n = 1e5,
  alpha_p = 0.2, beta_p = 10,
  t_min = -40,
  ms_epi = 0.1, sigma_epi = 0.5,
  ccf_thr_clade = 0.1, ccf_thr_count = 0.02,
  include_bp = T,
  n_chains = 4, 
  parallel_chains = 4,
  adapt_delta = 0.98,
  iter_warmup = 1000, iter_sampling = 5000
)
```

PEPI is performed with 4 parallel chains of 5000 samples and 1000 warmup iterations.

## Posterior and Prior

```{r driver-posterior}
pepi_obj <- PEPI::get_posterior(pepi_obj)
pepi_obj <- PEPI::get_prior(pepi_obj)
```
We collect prior and posterior from PEPI object.
## Epigenetic Switching Rate

```{r driver-s-epi}
p_omega <- PEPI::plot_inference(pepi_obj, parameter = c("omega_p"), log_scale = TRUE)

# Data frame for the line
vline_df <- data.frame(threshold = "Threshold", xval = 5e-3)

# Add geom_vline with aesthetic mapping
p_omega = p_omega +
  geom_vline(data = vline_df,
             aes(xintercept = xval, color = threshold),
             linetype = "dashed") +
  scale_color_manual(name = "Ground truth", values = "red")

p_s <- PEPI::plot_inference(
  pepi_obj,
  parameter = c("s_epi"), log_scale = F)
  
  # Data frame for the line
  vline_df <- data.frame(threshold = "Threshold", xval = 0.3)
  
  # Add geom_vline with aesthetic mapping
  p_s = p_s +
    geom_vline(
      data = vline_df,
      aes(xintercept = xval, color = threshold),
      linetype = "dashed"
    ) +
    scale_color_manual(name = "Ground truth", values = "red"
    )
  
  p_sd <- PEPI::plot_inference(
  pepi_obj,
  parameter = c("s_dc[1]"), log_scale = F)
  
  # Data frame for the line
  vline_df <- data.frame(threshold = "Threshold", xval = 0.25)
  
  # Add geom_vline with aesthetic mapping
  p_sd = p_sd +
    geom_vline(
      data = vline_df,
      aes(xintercept = xval, color = threshold),
      linetype = "dashed"
    ) +
    scale_color_manual(name = "Ground truth", values = "red"
    ) + xlim(0,1)

ggarrange(plotlist = list(p_omega,p_s,p_sd),nrow = 1,ncol = 3,common.legend = T)
```

We compare the ground truth with the inferred posterior distributions for $\omega_\oplus$, $s_\oplus$ and $s_\mathrm{dc}$


## Posterior Predictive Checks

```{r no_driver-ppc}
PEPI::plot_posterior_predictive(pepi_obj, parameter = "z_dc_driver[1,1,2]")
```

The predictive distribution aligns with observed number of mutation, indicating adequate model fit.


