---
title: "Model and notation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This page works through the mathematical model implemented by
`inst/multirates_positive_s.stan`, and lays each quantity next to the R
argument, table or column that supplies it. Read this before
`vignette("Multirates_inference")` if you want to know *why* the input
tables are shaped the way they are, rather than just how to build them.

## Two epigenetic states, growth and switching

Cells exist in one of two epigenetic states, written `-` and `+`. A
population's size in each state is a vector $Z(t) = (Z_-(t), Z_+(t))$,
and every lineage in the model (the wild-type background, a driver clade,
a driver+clade combo, ...) is governed by the same linear ODE, just with
its own rate parameters:

$$
\frac{dZ}{dt} = A\,Z, \qquad
A = \begin{pmatrix} \lambda_- & \omega_{+\to-} \\ \omega_{-\to+} & \lambda_+ \end{pmatrix}
$$

- $\lambda_-$ is the division rate of `-` cells.
- $\lambda_+ = \lambda_-(1+s)$ is the division rate of `+` cells, written
  relative to $\lambda_-$ via a fitness coefficient $s$ (so $s>0$ means
  `+` cells divide faster).
- $\omega_{-\to+}$ and $\omega_{+\to-}$ are the rates at which cells
  switch epigenetic state.

Because $A$ does not depend on $t$, the population at any later time is
an exact matrix-exponential solution:

$$
Z(t_2) = \exp\!\big(A\,(t_2-t_1)\big)\,Z(t_1)
$$

This is exactly `simulate_Z()`'s R implementation (via `expm::expm()`) and
the Stan model's own `Z()` Stan function - the same equation, evaluated
twice.

## The groups: wt, drivers, driver-only clades, driver+clade combos

The population is not a single lineage: it is the wild-type background
plus any number of sub-lineages, each founded at its own time from a
single cell and governed by the ODE above with its own rate parameters.

| Group | Founded at | Growth `s` | Switch rates | Confined to one state? |
|---|---|---|---|---|
| **wt** (background) | $t_{min}$ | $s_{epi}$ | $\omega_{n,wt}, \omega_{p,wt}$ | no |
| **clade_wt** (wt sub-lineage) | $t_{clade}\geq tmrca$ | $s_{epi}$ (shared with wt) | $\omega_{n,wt}, \omega_{p,wt}$ (shared with wt) | no |
| **driver** | $t_{driver}\geq tmrca$ | $s_{driver}$ (own) | $\omega_{n,driver}, \omega_{p,driver}$ (own) | no |
| **driver_n** | $t_{driver_n}\geq tmrca$ | $s_{driver_n}$ (own) | none (no switching) | yes, `-` only |
| **driver_p** | $t_{driver_p}\geq tmrca$ | $s_{driver_p}$ (own) | none (no switching) | yes, `+` only |
| **dc** driver leg | $t_{dc,1} = tmrca+\delta_{t,1}$ | $s_{dc}$ (own) | $\omega_{n,dc}, \omega_{p,dc}$ (own) | no |
| **dc** clade leg | $t_{dc,2} = t_{dc,1}+\delta_{t,2}$ | $s_{dc}$ (shared with its driver leg) | $\omega_{n,dc}, \omega_{p,dc}$ (shared) | no |

`driver_n`/`driver_p` are the simplest case: with no switching, the ODE
collapses to plain exponential growth, $Z(t) = \exp\big(\lambda_-(1+s)\,(t-t_{driver})\big)$.
A `dc` (driver+clade) entity is a *nested* pair: its driver leg is founded
first with a single cell, and its clade leg is a further sub-lineage
founded later *within* the driver leg, sharing the same growth/switch
rates - modelling a clade defined by a further epimutation switch event
that occurred after the driver mutation.

At each sampling time, the total population in each state is the sum of
the wt background and every additive group (`driver`, `driver_n`,
`driver_p`, and the `dc` driver leg - `clade_wt` and the `dc` clade leg
are *nested inside* their parent lineage's mass, not counted again):

$$
Z^{tot}_-(t) = Z_{wt,-}(t) + \sum_i Z_{driver_i,-}(t) + \sum_i Z_{driver\_n_i}(t) + \sum_i Z_{dc_i,-}(t)
$$

and analogously for $Z^{tot}_+(t)$.

## The observation model

### Cancer cell fractions (CCF) and mutation-arm fractions

A group's CCF at a sampling time and epistate is simply its share of the
total population in that epistate, $CCF_g(t) = Z_g(t)/Z^{tot}(t)$. This
is observed (as data) or predicted (as a posterior draw) with Beta noise
of concentration $\kappa$:

$$
CCF_g^{obs} \sim \text{Beta}\big(CCF_g \cdot \kappa,\ (1-CCF_g)\cdot\kappa\big)
$$

`kappa` is the same idea as `bayesplot`'s `neff_ratio` - it controls how
tightly the observed fraction is expected to hug its predicted mean;
high $\kappa$ means a very peaked, low-noise likelihood (see
`vignette("Convergence_diagnostics")` for why a very high `kappa` can make
Stan's sampler harder to initialize).

### Population sizes

Total population sizes in each state at sampling times (`zminus`,
`zplus`) and at any additional intermediate times (`ztot`) are observed
with lognormal noise of scale $\sigma_{count}$:

$$
z^{obs}(t) \sim \text{LogNormal}\big(\log z(t),\ \sigma_{count}\big)
$$

### Mutation counts

Every branch event (the trunk, a driver acquisition, a clade split, ...)
accumulates mutations as a Poisson process while cells divide. For a
branch with growth rate $\lambda$ (its own $\lambda_-$ or $\lambda_-(1+s)$
once a fitness effect has taken hold) spanning an elapsed time $\Delta t$:

$$
m \sim \text{Poisson}\big(4\,\mu\,\ell\,\lambda\,\Delta t\big)
$$

where $\mu$ (`mu`) is the per-division, per-bp mutation rate and $\ell$
(`l`) is the genome length. The trunk uses $\Delta t = tmrca - t_{min}$;
a driver or `driver_n` acquisition uses $\Delta t = t_{driver}-tmrca$; a
`dc` combo's clade leg uses the *post-driver* rate
$\lambda_-(1+s_{dc})$ over $\Delta t = t_{dc,2}-t_{dc,1}$.

## Priors

| Parameter | Prior | Set via |
|---|---|---|
| $\lambda_-$ (`lambda_n`) | $\text{Gamma}(\alpha_\lambda,\beta_\lambda)$ | `alpha_lambda`, `beta_lambda` |
| $s_{epi}$ (`s_epi`) | $\text{LogNormal}(ms_{epi},\sigma_{epi})$ | `ms_epi`, `sigma_epi` |
| $\omega_{n,wt},\omega_{p,wt}$ | $\text{Gamma}(\alpha,\beta)$ | `alpha_n_wt`/`beta_n_wt`, `alpha_p_wt`/`beta_p_wt` |
| $tmrca$ | $\text{Uniform}(t_{min}, t_{max})$ | `t_min` (`t_max` is the latest sampling time) |
| $s_{driver_i}, s_{driver\_n_i}, s_{driver\_p_i}, s_{dc_i}$ | $\text{LogNormal}(ms,\sigma)$, one prior per entity | `ms_driver`/`sigma_driver` columns (etc.) in the entity's `_muts` table |
| $\omega_{n,driver_i},\omega_{p,driver_i}$ (and `_dc`) | $\text{Gamma}(\alpha,\beta)$, one prior per entity | `alpha_n_driver`/`beta_n_driver` columns (etc.) |
| $\kappa$ | $\text{Uniform}(min_\kappa, max_\kappa)$ | `min_kappa`, `max_kappa` |
| $\sigma_{count}$ | $\text{Uniform}(min_{\sigma},max_{\sigma})$ | `min_sigma_count`, `max_sigma_count` |

Global priors are arguments to `fit_multirates()`. Per-entity priors
(driver/driver_n/driver_p/dc fitness and switch-rate hyperparameters) are
supplied per row in the corresponding `_muts` table, since a well-studied
oncogene and a novel candidate driver may legitimately warrant different
priors within the same fit.

## Notation table

The table below is the authoritative map from math to code: the Stan
parameter/generated-quantity name (as it appears in
`get_posterior_multirates()`'s tidy output), and the R-facing input that
supplies it.

```{r, echo = FALSE}
notation = data.frame(
  Symbol = c(
    "$\\lambda_-$", "$s_{epi}$ ($\\lambda_+=\\lambda_-(1+s_{epi})$)",
    "$\\omega_{-\\to+}$", "$\\omega_{+\\to-}$", "$tmrca$", "$t_{min}$",
    "$s_{driver_i}$", "$\\omega_{-\\to+,driver_i}$, $\\omega_{+\\to-,driver_i}$",
    "$t_{driver_i}$", "$m_{driver_i}$ (observed)",
    "$s_{driver\\_n_i}$ / $s_{driver\\_p_i}$", "$t_{driver\\_n_i}$ / $t_{driver\\_p_i}$",
    "$s_{dc_i}$", "$t_{dc_i,1}, t_{dc_i,2}$",
    "$t_{clade_i}$ (wt sub-lineage)",
    "$CCF_g(t)$ (observed)", "$\\kappa$",
    "$z_-(t), z_+(t)$ (observed)", "$\\sigma_{count}$",
    "$\\mu$ (mutation rate)", "$\\ell$ (genome length)"
  ),
  `Stan / getter name` = c(
    "`lambda_n`", "`s_epi`",
    "`omega_p_wt`", "`omega_n_wt`", "`tmrca`", "`t_min`",
    "`s_driver`", "`omega_p_driver`, `omega_n_driver`",
    "`t_driver`", "`m_driver`",
    "`s_driver_n` / `s_driver_p`", "`t_driver_n` / `t_driver_p`",
    "`s_dc`", "`delta_t_dc[,1]`, `delta_t_dc[,2]`",
    "`t_clade_wt`",
    "`ccf` (family `\"ccf\"`)", "`kappa`",
    "`zminus`, `zplus`", "`sigma_count`",
    "`mu`", "`l`"
  ),
  `R input` = c(
    "`fit_multirates(alpha_lambda=, beta_lambda=)`", "`fit_multirates(ms_epi=, sigma_epi=)`",
    "`fit_multirates(alpha_p_wt=, beta_p_wt=)`", "`fit_multirates(alpha_n_wt=, beta_n_wt=)`",
    "inferred (bounded by `t_min` and the latest sampling time)", "`fit_multirates(t_min=)`",
    "`driver_muts$ms_driver`, `sigma_driver`", "`driver_muts$alpha_n_driver`, ... (per row)",
    "inferred (bounded by `tmrca` and `t_max`)", "`driver_muts$m_driver`",
    "`driver_n_muts`/`driver_p_muts$ms_driver_n`/`_p`", "inferred",
    "`dc_muts$ms_dc`, `sigma_dc`", "inferred",
    "inferred",
    "`driver_ccf`/`clades_wt_ccf`/`dc_ccf`/`wt_ccf$ccf`", "`fit_multirates(min_kappa=, max_kappa=)`",
    "`population_sizes$zminus`, `zplus`", "`fit_multirates(min_sigma_count=, max_sigma_count=)`",
    "`fit_multirates(mu=)`", "`fit_multirates(l=)`"
  ),
  check.names = FALSE
)

knitr::kable(notation)
```

For the full input table schemas (columns required per group), see
`?init_multirates` and `?check_input_multirates`, or the worked example in
`vignette("Multirates_inference")`.
