---
title: "Multirates inference"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PEPI)
library(cmdstanr)
library(dplyr)
library(ggplot2)
```

PEPI's unified model (`inst/multirates_positive_s.stan`) jointly infers, from
multi-sample VAF-derived CCF data collected at several sampling times:

- the base growth rate `lambda_n` and epigenetic fitness `s_epi` of a
  wild-type ("wt") background population switching between "-" and "+"
  epistates at rates `omega_n_wt`/`omega_p_wt`;
- any number of **driver** clades (their own fitness `s_driver` and switch
  rates), **driver-only** clades confined to a single epistate
  (`driver_n`/`driver_p`), and **driver+clade** combos (`dc`, a driver
  acquisition followed by a further epimutation-switch event); and
- **wt-background clades** (`clades_wt`), sub-lineages of the wt population
  tracked for their own epistate composition.

## Simulate a synthetic dataset

We simulate one driver clade and one wt-background clade, both acquired
before the first sampling time, and observed at three time points. Data are
generated with the same deterministic growth/switch ODE the Stan model
itself assumes as its likelihood mean; `kappa` is kept modest here so the
simulated CCF/fraction noise is not razor-sharp, which keeps the model's
default initialization well-behaved for this small illustrative dataset.

```{r}

set.seed(1908)

sim = simulate_multirates_tree(
  sampling_times = c(3, 6, 9),
  tmrca = 1, t_min = 0,
  lambda_n = 1, s_epi = 0.15,
  omega_n_wt = 5e-3, omega_p_wt = 2e-3,
  drivers = list(
    list(id = "KRAS", t_driver = 2, s_driver = 0.3,
        omega_n_driver = 5e-3, omega_p_driver = 2e-3,
        ms_driver = -0.5, sigma_driver = 0.5,
        alpha_n_driver = 1, beta_n_driver = 10,
        alpha_p_driver = 1, beta_p_driver = 10)
  ),
  clades_wt = list(
    list(id = "c1", t_clade_wt = 1.5)
  ),
  mu = 1e-7, l = 2.7e9, kappa = 20, sigma_count = 0.1,
  seed = 42
)

sim$tables$population_sizes
sim$tables$driver_muts
sim$tables$driver_ccf
sim$truth

```

## Fit the model

```{r}

x = init_multirates(sim$tables, m_trunk = sim$truth$m_trunk)

x = fit_multirates(x, cmdstan_path = cmdstanr::cmdstan_path(),
                   method = "variational", ndraws = 1000, seed = 45,
                   mu = 1e-7, l = 2.7e9, t_min = 0,
                   ms_epi = 0, sigma_epi = 0.5,
                   alpha_lambda = 1, beta_lambda = 1,
                   alpha_n_wt = 1, beta_n_wt = 10, alpha_p_wt = 1, beta_p_wt = 10,
                   min_kappa = 5, max_kappa = 200,
                   min_sigma_count = 0.01, max_sigma_count = 1)

```

## Extract the posterior

```{r}

x = get_posterior_multirates(x)
x$posterior$multirates

```

## Predicted vs observed

```{r, fig.width=8, fig.height=6}
plot_counts_multirates(x)
```

```{r, fig.width=8, fig.height=6}
plot_fractions(x)
```

```{r, fig.width=8, fig.height=6}
plot_ccf(x)
```

```{r, fig.width=8, fig.height=5}
plot_mutations(x)
```

## Posterior vs prior, and ground-truth recovery

```{r, fig.width=8, fig.height=5}

plot_inference(x, params = c("lambda_n","s_epi","s_driver","omega_n_wt","omega_p_wt")) +
  geom_vline(
    data = data.frame(
      facet = c("lambda_n","s_epi","omega_n_wt","omega_p_wt","s_driver[KRAS]"),
      truth = c(sim$truth$lambda_n, sim$truth$s_epi, sim$truth$omega_n_wt,
               sim$truth$omega_p_wt, sim$truth$s_driver[["KRAS"]])
    ),
    aes(xintercept = truth), linetype = "dashed", color = "red"
  )

```

The dashed red lines mark the ground-truth values used to simulate the data;
posterior histograms should bracket them.
